Identifying Race Science in Mainstream Psychology, 1960-2010

### 2023-11-02 paper/paper.qmd
reformatting mostly finished

---
title: "Mainstreaming Race Science"
authors:
  - name: "Daniel J. Hicks"
    email: "dhicks4@ucmerced.edu"
    affiliations: 
      - id: ucm
        name: "University of California, Merced"
        city: "Merced"
        state: "CA, USA"
    corresponding: true
  - name: "Emilio J. C. Lobato"
    affiliations:
      - ref: ucm
abstract: |
  | **Background:** While anthropologists and human geneticists have long rejected the biological race concept, both social movements and prominent researchers have purported to justify racial inequality and colonialism by appealing to the epistemic authority of science. 
  | **Methods:** Bibliometric and text mining methods were used to trace the appearance of race science discourse in both mainstream scientific journals and the parascholarly journal *Mankind Quarterly* from 1960 to 2010. 
  | **Results:** Our analysis found two distinguishable lines of race science discourse.  One is strongly associated with *Mankind Quarterly*, while the other — emphasizing race and intelligence — is strongly associated with mainstream psychology journals, especially *Intelligence* and *Personality and Individual Differences*.  
  | **Conclusion:** Psychology, and especially intelligence research, has played a significant role in mainstreaming race science.

@@

bibliography: [references.bib, refs-manual.bib]
csl: nature.csl

@@

@@

| 
| **One sentence summary:** Bibliometric and text mining analysis indicate that race science discourse appeared in mainstream psychology journals throughout the second half of the twentieth century.

@@

We identified 16 researchers who had received funding from Pioneer; 14 of these researchers had profiles in the Web of Science (WoS) author search, allowing us identify 13 WoS-indexed journals that had published 6 or more of these authors. See @tbl-researchers and @fig-wos.

@@

@fig-wos shows that, while MQ is among the "Pioneer-publishing" journals, a number of mainstream journals are more prominent: *Personality and Individual Differences* (PID), *Intelligence* (Int), *Behavior Genetics* (BG), and *Psychological Reports* (PR).  In addition, only psychologist Richard Lynn appears to have published heavily in MQ.  Lynn became an assistant editor of MQ in 1979 (vol. XX, Nos. 1 & 2) and is listed as editor-in-chief on MQ's current website as of 2023-07-21.  He has also been president of Pioneer since the death of psychologist J. Phillippe Rushton in 2012 [@BeirichPioneerFundAssets2013].  By contrast, a number of Pioneer-funded researchers have published in PID, Int, and to a lesser degree BG: Bouchard, Eysenck, Jensen, Rushton, Vernon, and also Lynn.

@@

@fig-focal and this discussion focuses on three topics identified in one of the 24 fitted topic models; see figures @fig-gamma-lg, @fig-gamma-md, @fig-gamma-sm and supplemental visualizations S2-S4 for all topics from all models and S5-S7 for selected topics from all models.  In the focal model, topic 07 is strongly associated with MQ:  MQ published dozens of articles in this topic each year, and no other journal ever published more than a handful.  The "Silge plot," showing the top 15 terms in the topic [@SilgeTopicModeling2017], contains racial terms (`races`, `whites`, `negroes` and potentially `europe`, `africa`, `india`, and `japan`) as well as `book`, likely reflecting the fact that MQ published a number of book reviews, while the psychology journals either did not or these were not available for the corpus.

@@

Corpus assembly is summarized in @fig-corpus.  The corpus was assembled in two parts, one for *Mankind Quarterly* and the other for mainstream journals.

@@

After XML/PDF versions were retrieved, text was extracted using either the `xml2` package [@WickhamXml2ParseXML2023] or `pdftools`.  @fig-counts and @tbl-counts show the number of fulltext documents included in the corpus.  All together, the corpus includes 34,896 documents from 6 journals.

@@

After fitting topic models, our first research question was whether we could identify distinctive "race science discourse" topics.  [@fig-gamma-lg; @fig-gamma-md; and @fig-gamma-sm] show the gamma (topic-document) distributions for each value of $k$ for the three vocabularies.  In these figures, each panel corresponds to a single journal-$k$ combination, each row of cells is a single document in that journal, and each column of cells is a single topic in that particular model.  Topics correspond within columns of panels, but not within rows; for example, topic 04 for $k=10$ does not necessarily correspond to topic 04 for $k=20$.  Color intensity indicates the value of gamma for that particular topic-document combination.

@@

We next constructed Silge plots [@SilgeTopicModeling2017], showing the top 15 (highest-probability) terms (noun phrases) from each topic in the model.  These are provided as supplemental PDFs S2-S4.  Perusing these term lists, we focused on topics in each model that contained racial terms.  Every fitted topic model included at least one racial topic.  Importantly, in some cases these topics contained terms related to intelligence research, but in other cases they did not.

The gamma distributions and Silge plots suggested that, once $k$ was sufficiently large, the topic models were distinguishing between two types of race science articles.  We therefore examined the prevalence of topics that used racial terms, intelligence-research terms, or both, by journal and across all 24 models.  Supplemental PDFs S5-7 provide these visualizations.  In these visualizations, each row of facets corresponds to a value of $k$ (excluding 5 as too noisy), and each column of facets corresponds to one topic.  Topics are clustered based on whether the top 15 terms contain racial terms ("race"), terms related to intelligence research ("intelligence"), or both.  Line plots show the count of articles with $\gamma > 0.50$ for the given topic, by journal; thick lines are 5-year running averages and thin lines are raw counts.

@@

>>>>

---
title: "Identifying Race Science in Mainstream Psychology, 1960-2010"
authors:
  - name: "Daniel J. Hicks"
    email: "dhicks4@ucmerced.edu"
    affiliations: 
      - id: ucm
        name: "University of California, Merced"
        city: "Merced"
        country: "CA, USA"
    corresponding: true
  - name: "Emilio J. C. Lobato"
    affiliations:
      - ref: ucm
abstract: |
  | While anthropologists and human geneticists have long rejected the biological race concept, both social movements and prominent researchers have purported to justify racial inequality and colonialism by appealing to the epistemic authority of science. 
  Bibliometric and text mining methods were used to trace the appearance of race science discourse in both mainstream scientific journals and the parascholarly journal *Mankind Quarterly* from 1960 to 2010. 
  Our analysis found two distinguishable lines of race science discourse.  One is strongly associated with *Mankind Quarterly*, while the other — emphasizing race and intelligence — is strongly associated with mainstream psychology journals, especially *Intelligence* and *Personality and Individual Differences*.  
  Psychology, and especially intelligence research, has played a significant role in mainstreaming race science.

@@

bibliography: [references.bib, refs-manual.bib]
cite-method: "citeproc"
csl: "nature-no-superscript.csl"

@@

crossref:
  fig-prefix: ""

@@

<!-- NHB submission guidelines: <https://www.nature.com/nathumbehav/submission-guidelines> -->

@@

We identified 16 researchers who had received funding from Pioneer; 14 of these researchers had profiles in the Web of Science (WoS) author search, allowing us to identify 13 WoS-indexed journals that had published 6 or more of these authors. See @tbl-researchers and figure @fig-wos.

@@

Figure @fig-wos shows that, while MQ is among the "Pioneer-publishing" journals, a number of mainstream journals are more prominent: *Personality and Individual Differences* (PID), *Intelligence* (Int), *Behavior Genetics* (BG), and *Psychological Reports* (PR).  In addition, only psychologist Richard Lynn appears to have published heavily in MQ.  Lynn became an assistant editor of MQ in 1979 (vol. XX, Nos. 1 & 2) and is listed as editor-in-chief on MQ's current website as of 2023-07-21.  He has also been president of Pioneer since the death of psychologist J. Phillippe Rushton in 2012 [@BeirichPioneerFundAssets2013].  By contrast, a number of Pioneer-funded researchers have published in PID, Int, and to a lesser degree BG: Bouchard, Eysenck, Jensen, Rushton, Vernon, and also Lynn.

@@

Figure @fig-focal and this discussion focuses on three topics identified in one of the 24 fitted topic models; see figures [@fig-gamma-lg; @fig-gamma-md; @fig-gamma-sm; @fig-silge-lg; @fig-silge-md; @fig-silge-sm] for all topics from all models and figures [@fig-grid-lg; @fig-grid-md; @fig-grid-sm] for selected topics from all models.  In the focal model, topic 07 is strongly associated with MQ:  MQ published dozens of articles in this topic each year, and no other journal ever published more than a handful.  The "Silge plot," showing the top 15 terms in the topic [@SilgeTopicModeling2017], contains racial terms (`races`, `whites`, `negroes` and potentially `europe`, `africa`, `india`, and `japan`) as well as `book`, likely reflecting the fact that MQ published a number of book reviews, while the psychology journals either did not or these were not available for the corpus.

@@

Corpus assembly is summarized in figure @fig-corpus.  The corpus was assembled in two parts, one for *Mankind Quarterly* and the other for mainstream journals.

@@

After XML/PDF versions were retrieved, text was extracted using either the `xml2` package [@WickhamXml2ParseXML2023] or `pdftools`.  Figure @fig-counts and @tbl-counts show the number of fulltext documents included in the corpus.  All together, the corpus includes 34,896 documents from 6 journals.

@@

After fitting topic models, our first research question was whether we could identify distinctive "race science discourse" topics.  Figures [@fig-gamma-lg; @fig-gamma-md; and @fig-gamma-sm] show the gamma (topic-document) distributions for each value of $k$ for the three vocabularies.  In these figures, each panel corresponds to a single journal-$k$ combination, each row of cells is a single document in that journal, and each column of cells is a single topic in that particular model.  Topics correspond within columns of panels, but not within rows; for example, topic 04 for $k=10$ does not necessarily correspond to topic 04 for $k=20$.  Color intensity indicates the value of gamma for that particular topic-document combination.

@@

We next constructed Silge plots [@SilgeTopicModeling2017], showing the top 15 (highest-probability) terms (noun phrases) from each topic in the model; see figures [@fig-silge-lg; @fig-silge-md; @fig-silge-sm].  Perusing these term lists, we focused on topics in each model that contained racial terms.  Every fitted topic model included at least one racial topic.  Importantly, in some cases these topics contained terms related to intelligence research, but in other cases they did not.

The gamma distributions and Silge plots suggested that, once $k$ was sufficiently large, the topic models were distinguishing between two types of race science articles.  We therefore examined the prevalence of topics that used racial terms, intelligence-research terms, or both, by journal and across all 24 models; see figures [@fig-grid-lg; @fig-grid-md; @fig-grid-sm]. In these visualizations, each row of facets corresponds to a value of $k$ (excluding 5 as too noisy), and each column of facets corresponds to one topic.  Topics are clustered based on whether the top 15 terms contain racial terms ("race"), terms related to intelligence research ("intelligence"), or both.  Line plots show the count of articles with $\gamma > 0.50$ for the given topic, by journal; thick lines are 5-year running averages and thin lines are raw counts.

@@

::: {#refs}
:::

# Extended Data

\renewcommand{\thefigure}{ED-\arabic{figure}}
\setcounter{figure}{0}

![Silge plots (top 15 terms) for each topic and model, large vocabulary.](supplements/S2-lg-silge.pdf){#fig-silge-lg}

![Silge plots (top 15 terms) for each topic and model, medium vocabulary.](supplements/S3-md-silge.pdf){#fig-silge-md}

![Silge plots (top 15 terms) for each topic and model, small vocabulary.](supplements/S4-sm-silge.pdf){#fig-silge-sm}

![Topics with "race," "intelligence," or both occurring in the top 15 terms, large vocabulary.](supplements/S5_grid_lg.pdf){#fig-grid-lg}

![Topics with "race," "intelligence," or both occurring in the top 15 terms, medium vocabulary.](supplements/S6_grid_md.pdf){#fig-grid-md}

![Topics with "race," "intelligence," or both occurring in the top 15 terms, small vocabulary.](supplements/S7_grid_sm.pdf){#fig-grid-sm}
